Reasoning Graphically about Quantum Computation

### 2008-02-25 papers/reasoning-graphically/paper.tex
work
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1371 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

In graph patterns, variable nodes generalise the concept of a
boundary node. Variable nodes, unlike boundary-nodes in the concrete
graph, are not classified into domain and co-domain. The intuition of
a variable node is that it can be replaced by any concrete node. Thus,
unlike nodes in conrete a graph, variable nodes do not define an
ordering on the edges that are incident. Instead, the edges that touch
a variable node are simply a set.

The type of a variable nodes is thus defined as a multiset rather than
a tensor. Any tensor type can be lifted to a multiset type in the
obvious manner, which we denote by the operation {\tt
  mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor b
  \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
is an instance of a multiset type, $M$, when it's lifting into a
multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of a variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph. A variable node, $a$ is be said to an instance
of another variable node $b$ when the multiset representing the type of
$a$ is a superset of the multiset for the type of $b$. Thus, with
regard to the underlying semantics, $a$ being a subtype of $b$ means
that the set of instances of $a$ are a subset of the instances of $b$.

@@

@@

@@

@@

The purpose of our graph patterns is to express and formalise, in a
declarative manner, the intuitive idea of graph matching for graphs
that represent Compact Closed Categories. In the previous sections we
defined ways to extend the concrete graph representation with variable
nodes and !-boxes. In this section we combine these to first define
the matching relation between graph patterns and concrete graphs and
then introduce a matching relation between graph patterns.

\subsubsection{Concrete Matches}

We give a semantics to these in terms of a set of concrete graphs,
denoted by $\minterp{G}$, and say that a member of this set is
\emph{matched} by the graph pattern $G$. Formally, the set of matches
are defined as follows:

@@

where $\sinterp{G}$ represents the set of graphs that strictly contain
the graph $G$.

\subsubsection{Pattern Matches}

In this section, we define a compositonality principle for graphs
patterns, that is for graphs with variable nodes and !-boxes. This is
related to compositionality of the underlying concrete graphs by the
following instantiation homomorphism:

$\llbracket G \circ H \rrbracket = \llbracket G \rrbracket \circ \llbracket H \rrbracket$

By enriching graphs with to form
graph-patterns, we need to define compositionality principle from the
underlying concrete graphs. In this section, we restore
compositionality and show its correspondence to the underyling
principle in concrete graphs.

Definition: super-type: given a node with type T

Type Expressions: we introduce a new form of type for a pattern
graph. In particular, a variable node with

\subsection{Instantiation}

>>>>

In graph patterns, variable nodes generalise the concept of a boundary
node. Variable nodes, unlike boundary-nodes in the concrete graph, are
not classified into domain and co-domain. The intuition of a variable
node is that it can be replaced by any concrete node. Thus, unlike
nodes in conrete a graph, variable nodes do not define an ordering on
the edges that are incident. Instead, the edges that touch a variable
node are simply a set. The type of a variable nodes is thus defined as
a multiset rather than a tensor. Any tensor type can be lifted to a
multiset type in the obvious manner, which we denote by the operation
{\tt mset\_of\_tensor}. For example, {\tt mset\_of\_tensor($a \tensor
  b \tensor a$) = $\{(a,2), (b,1)\}$}. We say that a tensor-type, $T$,
is an instance of a multiset type, $M$, when it's lifting into a
multiset is a super-set of $M$: $ M \subseteq
\mathtt{mset\_of\_tensor}(T) \Longrightarrow T \in M$. The intuition
here is that an instance of a variable node is allowed to have as many
extra edges as it likes as long as it at least has the ones mentioned
in the original graph.

A variable node, $a$ is be said to an instance of another variable
node $b$ when the multiset representing the type of $a$ is a superset
of the multiset for the type of $b$. Thus, with regard to the
underlying semantics, $a$ being a subtype of $b$ means that the set of
instances of $a$ are a subset of the instances of $b$.

@@

A graph without boundary nodes but which may have variable nodes is
called a \emph{variable-node graph}. Such graphs are complete in the
sense that for every concrete graph there is a variable node graph
that can be instantiated to it: $\forall g.\,\exists G.\,g \in
\vinterp{G}$. The witness for this proof is made by simply replacing
all bounary nodes with variable nodes.

\subsection{Open Embedding}

@@

While concrete graphs can be composed sequentially as well as tensored
in an adjacent manner, variable-node graphs enjoy a single but
somewhat wilder joining operation which we call \emph{plugging}.

@@

\subsubsection{Making !-Box Graphs}

@@

The purpose of our graph patterns is to express and formalise, in a
declarative manner, the intuitive idea of graph matching for graphs
that represent Compact Closed Categories. In this section we combine
the two extensions to the concrete graph representation, namely
variable nodes and !-boxes, to provide a matching relation between
graph patterns and concrete graphs. We then introduce a matching
relation directly between graph patterns. The former relation gives a
semantics for our system and ensures soundness, while the latter
supports deirved equations and thus provides the theoretical basis for
our formulation of rewriting.

\subsubsection{Concrete Matches.}

We give a semantics for a pattern graph $G$ in terms of a set of
concrete graphs, denoted by $\minterp{G}$, and say that members of
this set are \emph{matched} by $G$. Formally, the set of matches are
defined as follows:

@@

\noindent where $\sinterp{G}$ represents the set of graphs that
strictly contain the graph $G$. Thus, a concrete graph $g$ is matched
by a graph pattern $G$, written $g \in G$, when an instance of the
pattern is a strict subgraph of the matched graph.

Graph patterns are complete in the sense that for every concrete graph
there is a pattern that matches it. This is a trivial consequence of
the completeness of variable-node extension to graphs, which is
unaffected by allowing !-boxes.

\subsubsection{Graph Pattern Matches.}

The definition of one pattern matching another one is simply that it
is more general. Formally, a graph pattern, $G$, matches another graph
pattern, $H$, when $\minterp{G} \supseteq \minterp{H}$. This provides
the specification for a matching algorithm which we provide an
implementation of in~\S\ref{sec:rewriting}.

Our representation has been carefully designed so that this concept
corresponds to what we call a \emph{respectful embedding} of $G$ into
$H$.

***

This is the basis for our implementation and is thus also the basis
for its correctness argument.
